---
title: "4. Generating GLOBIOM 30 region maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating GLOBIOM 30 region maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

For global analyses, GLOBIOM results are often presented for 30 major regions (ANYREGION). This vignette provides an example on how to put the results on a map. A world map that includes a mapping to the 30 GLOBIOM regions is included in the package and can be easily loaded (see below). 

## Creating the map

Plotting the world map with GLOBIOM regions is very similar to the procedure described to create the the [EU NUTS map](https://iiasa.github.io/globiomvis/articles/globiomvis3.html) so we will only provide a brief illustration here.

```{r message=FALSE, warning=FALSE, include=TRUE}
library(globiomvis)
library(ggplot2)
library(sf)
library(dplyr)

# Load the world map
world_poly <- st_read(system.file("shp", "region30.shp", package = "globiomvis"))

# We remove Antartica
world_poly <- world_poly %>%
  filter(iso3c != "ATA")

# As the palette has only 9 colors and we need 30, we need to interpolate additional colors.
library(RColorBrewer)
n_col = length(unique(world_poly$ANYREGION))
getPalette = colorRampPalette(brewer.pal(12, "Paired"))

# Create the world map showing the GLOBIOM regions.
# We add a grey background map that will show polygons not included in ANYREGION.
ggplot() +
  geom_sf(data = world_poly, colour = "grey30", fill = "grey80") +
  geom_sf(data = filter(world_poly, !is.na(ANYREGION)), aes(fill = ANYREGION)) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(fill = NULL) +
  scale_fill_manual(values = getPalette(n_col))
```


