---
title: "1. Install R, packages and other relevant software"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Install R, packages and other relevant software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing R and RStudio

To use the package one needs to install R and RStudio. R can be downloaded [here](https://cran.r-project.org/bin/windows/base/) and is available for various operating systems. I recommend not to use the latest release as it takes some time before all packages are updated and therefore might results in errors. I normally use the second to last release.

Rstudio is most used open source software to code in R (but it can do a lot more). It can be downloaded [here](https://rstudio.com/products/rstudio/download/). The free version is sufficient. 


## Installing GDXRRW

To interface between R and GAMS the GDXRRW package is required. This package is not available o CRAN, the R package repository, and therefore cannot be installed using the standard approach (see next section). Hence, one has to install it manually. The package can be found [here](https://support.gams.com/gdxrrw:interfacing_gams_and_r)

To install GDXRRW:

1. Copy the latest version (GDXRRW.zip) to a temporary folder on your machine
2. Click 'Packages' > 'install' in RStudio
3. Select 'Install from: Package Archive File (.zip; .tar; .gz)
4. Select GDXRRW.zip to install the package in your R library

Note that the version of GDXRRW has to match your R version, otherwise there might be an error. Thus, if you have R version 3.5.X install the GDXRRW version that is built with R 3.5.x.


## Installing R packages

In contrast to GDXRRW, most packages can be easily installed from CRAN by using the following command: `install.packages()`). In order to install rglobiom, which is also not on CRAN, one needs the package devtools. This is installed in the following way:

```{r, eval=FALSE}
install.packages("devtools")
```

A number of other packages, most importantly dplyr, purrr, tidyr and ggplot2, which are all part of the [tidyverse](https://www.tidyverse.org/) and are used by rglobiom, will be installed automatically when rglobiom is installed. 


## Installing rglobiom

To install rglobiom, first install the devtools package and do the following:

```{r, eval=FALSE}
library(devtools)
devtools::install_github("iiasa/rglobiom")
```


## Additional software

Although not strictly necessary to use rglobiom (for now), I recommend installing the following additional software:

- osgeo4w. This is a collection of open source geospatial software for windows and can be downloaded [here](https://trac.osgeo.org/osgeo4w). Most important part is gdal, which is a very powerful translator library for raster and vector geospatial data formats. It is used by several R packages that process raster and vector data and is normally much faster than processing this data within R. osgeo4w is only available for windows. gdal for other systems can be downloaded [here](https://gdal.org/download.html).

- miktex. This is (one of) implementations of TeX/LaTeX and related programs. RStudio facilitates the use of LaTex through Rmarkdown but this requires the installation of [miktex](https://miktex.org). If you want to learn more about this look [here](https://bookdown.org/yihui/rmarkdown/)

- git. This is software to version you code and can be downloaded [here](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git). RStudio and git can be easily linked to version your code and use [github](https://github.com) to store code in the cloud and facilitate collaborative code development. More information about how to set this up can be found [here](https://www.r-bloggers.com/rstudio-and-github/) and [here](https://www.r-bloggers.com/using-rstudio-and-git-version-control/).
