---
title: "Install R, packages and other relevant software"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Install R, packages and other relevant software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing R and RStudio

Since `globiomvis` is an R package, R is required. R can be downloaded [here](https://cran.r-project.org). On Linux, you can instead install R with the package manager of your distribution.

We highly recommend to also install the RStudio integrated development environment. It can be downloaded [here](https://rstudio.com/products/rstudio/download/). It will allow you to easily try the examples and debug your code. Some of the discussion that follows assumes that you have RStudio installed.

## A note on R packages

Packages are bundles that contain R scripts and sometimes code written in other languages (e.g. C or C++) for speedings up calculations or interfacing with other software (e.g. GAMS). Packages can easily be installed by typing `install.packages("<package name>")` or by clicking on the 'package' tab in RStudio. Most packages can be downloaded from CRAN, the official R package repository.

The `globiomvis` package depends on other R packages. A large subset of these are jointly referred to as the [tidyverse](https://www.tidyverse.org/), a curated package collection that includes `dplyr`, `purrr`, `tidyr` and `ggplot2`.

Installing packages sometimes results in conflicts when an older version of these packages is already installed on your system and has been loaded into memory. To prevent this, install packages immediately after (re)starting R. If you get the message `These package have more recent versions available. It is recommended to update all of them. Which would you like to update?`, the conservative choice is  to select `3: None`: you can always update later after further deliberation.

## Installing globiomvis


From an R or RStudio console, installing the `globiomvis` package can be as simple as:

```{r, eval=FALSE}
# Install the devtools package, if you have not done so already
install.packages("devtools")
# Install globiomvis development version from GitHub, it is not available on CRAN
devtools::install_github("iiasa/globiomvis")
```


This should also install all further required packages, but the `gdxrrw` package will fail to install if your system lacks a compiler. You can either install the required compiler tools, which will add a lot of stuff to your system, or you can install a binary version of the `gdxrrw` package as documented in the next section.

## Installing a gdxrrw binary package

To read/write GDX files from R, the `gdxrrw` package is required. This package is not available on CRAN and therefore cannot be installed using the standard approach (see next section). The package can be found [here](https://support.gams.com/gdxrrw:interfacing_gams_and_r)

To install `gdxrrw` without compiling, download the latest binary package built for your operating system and R version (if you have R version 3.5.X install the `gdxrrw` version that is built with R 3.5.x) and save the package archive file to a tempory directory. Then:

1. Click 'Packages' > 'install' in RStudio
2. Select 'Install from: Package Archive File (.zip; .tar; .gz)'
3. Select the package archive file in the temporary directory to install the package in your R library

## Installing compiler tools

Linux should already have a compiler installed. On Windows, compiling code in source-only packages requires an [Rtools](https://cran.r-project.org/bin/windows/Rtools) installation. On MacOS, the [development tools and libraries](https://cran.r-project.org/bin/macosx/tools) are required for compilaton.

After having installed these tools, performing the `globiomvis` installation as explained above should succeed in installing `gdxrrw`.

## Using gdxrrw

For `gdxrrw` to work, the package needs to load GDX libraries from the GAMS system directory. Use the `igdx()` function to tell `gdxrrw` where it can find GAMS. There are multiple ways to do so as explained in the `igdx` help page:

```{r, eval=FALSE}
library(gdxrrw)
help(igdx)
```

You can specify the GAMS system directory explicitely. For example:

```{r, eval=FALSE}
igdx("C:/GAMS/win64/29.1")
```

However, we recommended to set the `R_GAMS_SYSDIR` environment variable. For details see [here](https://iiasa.github.io/GLOBIOM/R.html). After setting `R_GAMS_SYSDIR`, you should be able to load the GDX libraries by passing an empty string as first argument to `igdx`:

```{r, eval=FALSE}
igdx("")
```

In the examples that follow, it is assumed that you have set `R_GAMS_SYSDIR`.

## Additional software

Although not strictly necessary to use `globiomvis` (for now), we recommend installing the following additional software:

* **GDAL**. (When not on Windows). [GDAL](https://en.wikipedia.org/wiki/GDAL) is a very powerful translator library for raster and vector geospatial data formats. It is used by several R packages that process raster and vector data and is normally much faster than processing this data within R. GDAL can be downloaded [here](https://gdal.org/download.html). Alternatively, on Linux, the packages `libgdal-dev` and `libudunits2-dev` can be installed.

* **OSGeo4W**. (Windows) This is a collection of open source geospatial software for Windows that provides GDAL and much more. It can be downloaded [here](https://trac.osgeo.org/osgeo4w).

* **MiKTeX** is one of the distributions of TeX/LaTeX and related programs. RStudio leverages LaTeX through [R Markdown](https://rmarkdown.rstudio.com/docs/index.html) by, for example, rendering PDF documents from R Markdown files via LaTeX. To learn more about this see [here](https://bookdown.org/yihui/rmarkdown/).

  Unfortunately, the basic MikTeX installer does not install everything that is required. Instead, from the 'All downloads' tab on the [MiKTeX download page](https://miktex.org/download#all), download the Net Installer and use it to perform a complete installation. This will require roughly 7GB of disk space.

* **TinyTeX**. If you [prefer a smaller](https://github.com/yihui/tinytex/issues/7) TeX/LaTeX installation, you can try [TinyTeX](https://yihui.org/tinytex/) instead. It has been designed to work well with R. Beware though that TinyTeX is still maturing. 

* **Git** version control software. It can be downloaded [here](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git). RStudio and Git can be easily linked to version your code. Git repositories can be hosted on [GitHub](https://github.com/about), a cloud platform that facilitates collaborative code development. More information about how to set this up can be found [here](https://www.r-bloggers.com/rstudio-and-github/) and [here](https://www.r-bloggers.com/using-rstudio-and-git-version-control/).
