---
title: "2. Generating GLOBIOM scenario plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating GLOBIOM scenario plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

To quikly inspect GLOBIOM runs it is often useful to compare historical trends of key variables with model projections. The package includes several functions to support this task. `trend_plot_all` produces a pdf file with plots for key combinations of variables, items and units across all GLOBIOM regions for which data is created. `trend_plot` is basic function that allows the user to make individual plots for selected variables, items, units and regions. Before the functions can be used historical data and globiom output needs to be loaded first. 


## Loading and preparing data

Start by loading the required packages. Apart from the gdxrwww package, all packages can be installed from CRAN, the R package repository. See the Vignette '1. Install R and other software' for instructions if you do not know how to do this.

```{r message=FALSE, warning=FALSE, include=TRUE}
library(gdxrrw)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(rglobiom)
```

Next, we need to link R and GAMS so that R is able to load GDX files using the command `igdx()`. Use the folder where gams.exe is stored on your machine. If the link is created you should get the message `The GDX library has been loaded`.

```{r}
gams_path <- "c:/GAMS/win64/25.0"
igdx(gams_path)
```

We load a GLOBIOM output file. We use `setNames` to add column names. The will be missing if the gams data table (i.e. the Symbol) has '*' instead of variables. We also ensure that `YEAR` is an integer (and not a character) and `ITEM_AG, VAR_ID and VAR_UNIT` are in capitals, and add a 'source' column.

```{r eval=FALSE}
globiom_path <- "P:/globiom"
globiom_file <- file.path(globiom_path, "projects/TEMP/globiom_results/MODEL_OUTPUT_AG_SSP2_lookup_q_SDG_10052019.gdx")
symbol <- "OUTPUT_AG"
globiom <- rgdx.param(globiom_file, symbol) %>%
  setNames(c("VAR_ID", "VAR_UNIT", "REGION_AG", "ITEM_AG", "MacroScen", "BioenScen", "IEA_Scen", "YEAR", "OUTPUT_AG")) %>%
  mutate(YEAR = as.integer(as.character(YEAR)),
         ITEM_AG = toupper(as.character(ITEM_AG)),
         VAR_ID = toupper(VAR_ID),
         VAR_UNIT = toupper(VAR_UNIT),
         source = "globiom") %>%
  droplevels
```

We also load the historical FAOSTAT data, aggregated to GLOBIOM nomenclature and make some adjustments in the same way as we did for the globiom data file.

```{r eval=FALSE}
hist <- rgdx.param(file.path(globiom_path, "projects/TEMP/globiom_results/OUTPUT_FAO_REGION_since1961_ag_NSUST.gdx"), "OUTPUT_AG") %>%
  mutate(YEAR = as.integer(as.character(YEAR)),
         ITEM_AG = toupper(as.character(ITEM_AG)),
         VAR_ID = toupper(VAR_ID),
         VAR_UNIT = toupper(VAR_UNIT),
         source = "historical") %>%
  droplevels
```


## Create a pdf with plots for key variable, item and unit combinations

Main combinations of globiom variables (VAR_ID), items (ITEM_AG) and units (VAR_UNIT) are stored in `main_output_comb`. The function `trend_plot_all` creates a pdf file with plots for all of these output combinations (and are present in the GLOBIOM output file). In the example below the pdf file with the name `globiom_trend_plots_YYYY-MM_DD.pdf` is saved to the folder `c:/globiom`, which is created first. Note that it might take a bit of time as more than 200 plots are generated.

```{r eval=FALSE}
dir.create("c:/globiom", showWarnings = F)
trend_plot_all(df_gl = globiom, df_hs = hist, path = "c:/globiom")
```

## Create plots for your own selection of GLOBIOM output

It is also possible to create a pdf with a subset of combinations. The function `all_output_comb()` creates a data frame with all VAR_ID, ITEM_AG and VAR_UNIT combinations in a globiom output file. This file can be used as a basis to filter out relevant output combinations. Alternatively, one can use `main_output_comb` as a basis or create a new data frame with output combinations. 

```{r eval=FALSE}
comb_sel <- all_output_comb(globiom) %>%
  filter(VAR_ID %in% c("AREA", "PROD"), ITEM_AG %in% c("CORN", "CRPLND"), VAR_UNIT %in% c("1000 HA", "1000 T"))

trend_plot_all(df_gl = globiom, df_hs = hist, path = "c:/globiom", comb = comb_sel)
```

If you want full flexibility to plot the GLOBIOM results, for instance by zooming on a subset of scenarios for a selected number of variables, items and units, it is easier to use the basic function `trend_plot`. The following example uses `trend_plot` to create a plot for VAR_ID = AREA, ITEM_AG = 1000 HA, VAR_UNIT = 1000 HA, regions: World and EUR and two scenarios: GHG000_BIO00 and GHG000_BIO06.

```{r eval=FALSE}
trend_plot(var = "AREA", item = "CORN", unit = "1000 HA", reg = c("World", "EUR"), df_gl = globiom, df_hs = hist)
```

If you want multiple figures, one has to loop this function over a selection of the data. This requires some additional code and is illustrated below. 

```{r eval=FALSE}
# Select scenarios, variables, items and units
var_sel <- c("AREA", "PROD")
item_sel <- c("CORN", "CRPLND")
unit_sel <- c("1000 HA", "1000 T")
reg_sel <- c("World", "EUR")
scen_sel <- c("GHG000_BIO00", "GHG000_BIO06")

# Create data frame with selected GLOBIOM output
globiom_sel <- globiom %>%
  filter(VAR_ID %in% var_sel , ITEM_AG %in% item_sel , VAR_UNIT %in% unit_sel, BioenScen %in% scen_sel)

# Create data frame with unique output combinations in selected globiom output
sel_output_comb <- all_output_comb(globiom_sel)

# Loop over VAR_ID, ITEM_AG and VAR_UNIT 
pdf(file = file.path("c:/globiom", paste0("globiom_trend_plots_", Sys.Date(), ".pdf")))
    purrr::walk(1:nrow(sel_output_comb), function(i){
      trend_plot(var = sel_output_comb$VAR_ID[i], item = sel_output_comb$ITEM_AG[i], 
                 unit = sel_output_comb$VAR_UNIT[i], reg = reg_sel, df_gl = globiom_sel, df_hs = hist)
    })
    dev.off()
    
```


